Demystifying the Computer Science Ph.D. Admission in the US \\{\Large A guide for Vietnamese and International Students

### 2023-06-10 phd-cs-us.tex
Updates from Overleaf

\newtcolorbox{cbox}[]
{
left=1pt,right=1pt,top=1pt,bottom=1pt,
colback=gray!10,
boxrule=0.5pt
}

@@

\newcommand{\mycomment}[3][\color{black}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}
\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\title{Demystifying the Computer Science Ph.D. Admission in the US}
\date{\today}
\author{ThanhVu (Vu) Nguyen, George Mason University}
%\setcounter{secnumdepth}{3}
\begin{document}
\maketitle

@@

So I want to share my opinions and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While I wrote this document for Vietnamese students interested in CS, it should be applicable to students in various countries interested in STEM.
Moreover, while many examples given are for GMU, which is 33rd on CSRankings (see \S\ref{sec:ranking}), the writing should be generalize to most other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines.} universities  (though \emph{very} top schools might be very extreme, e.g., see the admission process at \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{CMU}).

@@

\begin{commentbox}
Vu: We usually decide that a PhD candidate is either (i) admit with funding (TA or RA) or (ii) rejected. In other words, in most cases, we either
admit you with full funding, or we don't. In some rare cases, we may admit
you with no funding because you have funding on your own (e.g., you are
supported by your government or have external grants). We justify
our decision with a short summary about your application, where we list
strengths (e.g., well-known undergrad school) and weaknesses (e.g., weak
LORs).
\end{commentbox}

@@

\paragraph{English Test} Unless you're from certain countries (e.g., English speaking ones), you will need to
take standardized English test. Just do well enough to pass minimum requirement set by the university, which nowadays has many options for you to choose from.

\begin{commentbox}
Vu: The minimum for GMU (being above this might not mean much, but below is a \red{red flag}).
\begin{itemize}
\item GPA: $\ge 3.0$ in your undergrad (but as mentioned, we also consider the rank/prestige of your school)
\item GRE: not required, though it can help boost your profile
%    \item but if you want to use it, then we expect a total (V+Q) of $\ge 311$ (with a $\ge 157$ Q) and A $\ge 3.0-3.5$.
\item English requirement tests (one of the below)
  \begin{itemize}    
  \item TOEF: 88 pts in total AND $\ge 20$ points in each subsection OR
  \item IELTS: $\ge 6.5$ OR
  \item DuoLingo Graduate English: $\ge 120$ OR 
  \item Pearson Test of Academic English: $\ge 67$
  \end{itemize}  
\end{itemize}
\end{commentbox}

@@

Fellowship is another type of funding in which the student applies for (e.g., from school, industries, government). Fellowships are typically competitive and generous, and gives pretty much all benefits tuition/insurance that a TA/RA has.  Moreover, it often gives higher stipend (including summer) and opens doors for job opportunities (e.g., internship).  For example, a student with a Microsoft fellowship will likely get an internship at Microsoft.

@@

\textbf{How to get Fellowship?} You apply for them.  The US government has many fellowships but these would likely require US citizenship or residency.  However, tech companies including Google, Microsoft, Facebook, IBM have fellowships that international students can apply for.

@@

\subsection{Current 2023 Rankings of CS PhD programs in the U.S}\label{sec:ranking}
  Tab.~\ref{tab:ranking} lists the top 50 CS programs in the US from \href{https://www.csrankings.org}{CSRankings.org}, a ranking system that is based on CS conferences.

  \begin{table}
    \centering
    \small
    \caption{Top 50 CS PhD programs in the U.S. (CSRankings, April 2023)}\label{tab:ranking}
  \begin{tabular}{rl|rl}
    \toprule
    1 & Carnegie Mellon University & 26 & Univ. of California - Irvine \\
    2 & Univ. of Illinois at Urbana-Champaign  & 27 & University of Southern California \\
    3 & Univ. of California - San Diego & 28 &  Duke University \\
    4 & Massachusetts Institute of Technology & 29 & Univ. of California - Riverside\\
    5 & Stanford University         & 30 & Pennsylvania State University \\
    6 & University of Michigan - Ann Arbor & 31 & Northwestern University \\
    7 & University of Washington   & 32& Ohio State University \\  
    8 & Cornell University       &33 & George Mason University \\
    9 & Univ. of California - Berkeley  &34& Harvard University  \\
    10 & Georgia Institute of Technology &35& Brown University  \\
    11 & Northeastern University  &36& University of Utah \\
    12 & University of Maryland - College Park &37& Texas A\&M University\\
    13 & University of Wisconsin - Madison  &38& Univ. of California - Santa Cruz  \\
    14 &   Purdue University  &39 & Yale University  \\
    15 & University of Texas at Austin  &40& Boston University \\
    16 &   Columbia University  &41 & University at Buffalo\\
    17 & University of Pennsylvania &42& University of Colorado-Boulder \\
    18 & New York University  & 43& North Carolina State University\\
    19 & Princeton University & 44& Rice University \\
    20 & University of Massachusetts Amherst &45& University of Illinois at Chicago  \\
    21 & Univ. of California - Los Angeles &46& University of North Carolina \\
    22 & University of Chicago &47& University of Virginia \\
    23 & Stony Brook University &48& Arizona State University \\
    24 & Univ. of California - Santa Barbara &49& University of Minnesota \\
    25 & Rutgers University &50& Virginia Tech  \\
    \bottomrule
    \end{tabular}
\end{table}

>>>>

\newtcolorbox{cbox}
{
left=1pt,right=1pt,top=1pt,bottom=1pt,
colback=gray!10,
boxrule=0.5pt
}

@@

\newcommand{\mycomment}[3][\color{black}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}
\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\title{Demystifying the Computer Science Ph.D. Admission in the US \\{\Large A guide for Vietnamese and International Students}}
\date{\today}
\author{\small ThanhVu (Vu) Nguyen, George Mason University}
%\setcounter{secnumdepth}{3}
\begin{document}
\maketitle

@@

So I want to share my opinions and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While I wrote this document for Vietnamese students interested in CS, it should be applicable to students in various countries interested in STEM.
Moreover, while many examples given are for GMU, which is 32rd in the US according CSRankings (see \S\ref{sec:ranking}), the writing should be generalize to most other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines.} universities  (though \emph{very} top schools might be very extreme, e.g., see the admission process at \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{CMU}).

@@

\begin{commentbox}
Vu: We usually decide that a PhD candidate is either (i) admit with funding (TA or RA) or (ii) rejected. In other words, in most cases, we either
admit you with full funding, or we don't. In some rare cases, we may admit
you without funding because you have funding on your own (e.g., you are
supported by your government or have external grants). We also justify
our decision with a summary about your application, where we list
strengths (e.g., well-known school) and weaknesses (e.g., weak
LORs).
\end{commentbox}

@@

\paragraph{English Test} Unless your degrees are from certain countries such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{these}, you will need to
take standardized English test. Just do well enough to pass minimum requirement set by the university, which nowadays has many options for you to choose from.

\begin{commentbox}
Vu: The minimum for GMU (being above this might not mean much, but below is a \red{red flag}).
\begin{itemize}
\item GPA: $\ge 3.0$ in your undergrad (but we also consider the rank/prestige of your school)
\item GRE: not required, though it can help boost your profile
%    \item but if you want to use it, then we expect a total (V+Q) of $\ge 311$ (with a $\ge 157$ Q) and A $\ge 3.0-3.5$.
\item English requirement tests (one of the below)
  \begin{itemize}    
  \item TOEF: 88 pts in total AND $\ge 20$ points in each subsection OR
  \item IELTS: $\ge 6.5$ OR
  \item DuoLingo Graduate English: $\ge 120$ OR 
  \item Pearson Test of Academic English: $\ge 67$
  \end{itemize}  
\end{itemize}
\end{commentbox}

@@

Fellowship is another type of funding that the student needs to apply for (e.g., from school, industries, government). Fellowships are typically competitive and generous, and gives pretty much all benefits tuition/insurance that a TA/RA has.  Moreover, it often gives higher stipend (including summer) and opens doors for job opportunities (e.g., internship).  For example, a student with a Microsoft fellowship will likely get an internship at Microsoft.

@@

\textbf{How to get Fellowship?} You apply for them.  The US government has many fellowships that would likely require US citizenship or residency.  However, tech companies including Google, Microsoft, Facebook, IBM have fellowships that international students can apply for.

@@

\subsection{Current 2023 Rankings of CS PhD programs in the U.S}\label{sec:ranking}
  Tab.~\ref{tab:ranking} lists the top 50 CS programs in the US from \href{https://www.csrankings.org}{CSRankings.org}, a ranking system  based on top CS conferences.

  \begin{table}[h!]
    \centering
    \small
    \caption{Top 50 CS PhD programs in the U.S. (CSRankings, June 2023)}\label{tab:ranking}
  \begin{tabular}{rl|rl}
    \toprule
    1 & Carnegie Mellon & 26 & Univ. of California - Irvine \\
    2 & Univ. of Illinois at Urbana-Champaign  & 27 &  Duke University \\
    3 & Univ. of California-San Diego & 28 & Rutgers University \\
    4 & MIT & 29 & Univ. of California - Riverside\\
    5 & Georgia Institute of Technology         & 30 & Northwestern University\\
    6 & Stanford University& 31 & Pennsylvania State University  \\
    7 & University of Michigan - Ann Arbor   & 32& George Mason University\\  
    8 & University of Washington      &33 &  Harvard University \\
    9 &  Univ. of California - Berkeley  &34&  Univ. of California - Santa Cruz \\
    10 & Cornell University  & 35 &  Yale University \\
    11 & University of Maryland - College Park &  36& Brown University \\ 
    12 & Northeastern University 37&  Ohio State University\\
    13 & University of Wisconsin - Madison  &38& Texas A\&M University \\
    14 & Columbia University   &39 & Boston University  \\
    15 &   Purdue University  &40& North Carolina State University\\\  
    16 & University of Texas at Austin   &41 & University of Utah \\
    17 & University of Pennsylvania &42 & University at Buffalo\\
    18 & Princeton University  & 43& Rice University\\
    19 & University of Massachusetts-Amherst & 44&  University of Colorado-Boulder \\
    20 &  New York University  &45& University of Illinois at Chicago  \\
    21 & Univ. of California - Los Angeles &46& Virginia Tech  \\
    22 & University of Southern California &47&  Arizona State University \\
    23 & Stony Brook University &48&University of Minnesota \\
    24 & University of Chicago &49& University of Virginia \\
    25 & Univ. of California - Santa Barbara &50& University of North Carolina \\
    \bottomrule
    \end{tabular}
\end{table}
